---
title: "Early C2C12 Growth"
author: "Dr. Sara J. Smith"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: yes
    latex_engine: xelatex
  html_document:
    df_print: kable
    toc: yes
    toc_float: yes
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 80))
```

## **Load Libraries & Data**

Load necessary libraries and data, clean and refactor data.

```{r, message = F, warning = F}
library(readxl)
library(lmerTest)
library(DHARMa)
library(emmeans)
library(fitdistrplus)
library(PNWColors)
library(tidyverse)
```

```{r, echo = F, message = F, warning = F}
myo <- read_excel("serum.xlsx", sheet = 1) %>% 
  mutate(condition = fct_relevel(as.factor(condition), "10FBS"),
         trial = as.factor(trial), 
         day = fct_relevel(as.factor(day), "MB"))

p21 <- read_excel("serum.xlsx", sheet = 2) %>% 
  mutate(condition = fct_relevel(as.factor(condition), "10FBS"),
         trial = as.factor(trial), 
         day = fct_relevel(as.factor(day), "MB"))

cfi <- read_excel("serum.xlsx", sheet = 3) %>% 
  mutate(condition = fct_relevel(as.factor(condition), "10FBS"),
         trial = as.factor(trial)) %>%
  select(-c(day))

diam <- read_excel("serum.xlsx", sheet = 4) %>% 
  mutate(condition = fct_relevel(as.factor(condition), "10FBS"),
         trial = as.factor(trial), 
         day = fct_relevel(as.factor(day), "7"))

diam.clean <- diam %>%
  group_by(condition, day, trial) %>%
  summarize(avg.diam = mean(diameter))
```

## **Myogenin Expression**

### *Modelling*
```{r}
myo.mod <- lmer(deltacq ~ condition*day + (1|trial), data = myo)
myo.res <- simulateResiduals(myo.mod)
plot(myo.res)
```

```{r}
summary(myo.mod)
```

- Intercept: Represents the predicted deltacq value for the 10FBS
  condition at Day MB

- Condition effects (at Day MB): Compared to 10FBS at the MB time point,
  `deltacq` is significantly higher across all other media

- Day effects (in 10FBS): Relative to Day MB, `deltacq` significantly
  decreases over time within the baseline 10FBS condition

- Condition \* Day interactions: None of the `condition:day1`
  interaction terms are strong. This indicates that the change in
  `deltacq` from Day MB to Day 1 in 10HS, 2FBS, and 2HS is not
  meaningfully different from the drop observed in 10FBS.

- All condition:day2 terms were automatically dropped because of the
  model matrix rank deficiency. Model is built in order of Intercept,
  Main Effects, then Interactions, which "uses up" all available degrees
  of freedom before getting to day 2 in interaction terms

Let's try a model without MB to capture a full factorial design:

```{r}
myo.diff <- myo %>%
  filter(day == "1" | day == "2")
  
myo.diff.mod <- lmer(deltacq ~ condition * day + (1 | trial), data = myo.diff)
myo.diff.res <- simulateResiduals(myo.diff.mod)
plot(myo.diff.res)
```

```{r}
summary(myo.diff.mod)
```

- Intercept: Represents the baseline predicted deltacq for 10FBS on Day 1.  

- Day 1 Condition effects: On Day 1, all test conditions show
  significantly higher `deltacq` values compared to the baseline 10FBS

- Time (day 1 to day 2): In the 10FBS baseline condition, `deltacq`
  significantly decreases by 2.13 units from Day 1 to Day 2.

- Condition \* Day interactions: None of the interaction terms exhibit
  differences. This indicates that the rate of change from Day 1 to Day
  2 in 10HS, 2FBS, and 2HS does not meaningfully differ from 10FBS.


#### *New model Sept 14 2026*

The interaction term in the previous model (`myo.diff.mod`) was not significant, and therefore we can drop the interaction term, model just fixed effects and extract estimated marginal means from that model instead. 

```{r}
myo.diff.mod2 <- lmer(deltacq ~ condition + day + (1 | trial), data = myo.diff)
myo.diff.res2 <- simulateResiduals(myo.diff.mod2)
plot(myo.diff.res2)
```

```{r}
AIC(myo.diff.mod, myo.diff.mod2) # effectively equal
summary(myo.diff.mod2)
```

Extracting estimated marginal means
```{r}
myo.cond.emmeans <- emmeans(myo.diff.mod2, pairwise ~ condition | day)
myo.cond.emmeans
```

```{r}
myo.day.emmeans <- emmeans(myo.diff.mod2, pairwise ~ day | condition)
myo.day.emmeans
```


### *Visualization*

Colour palette for MB, Day 1, Day 2
NB: can convert all colour palettes to a colourblind or grey scale friendly palette
```{r}
day.cols <- pnw_palette(name = "Starfish", n = 3, type = "discrete")
```
  
Myogenin expression
```{r}
ggplot(myo, aes(x = day, y = deltacq, fill = day, colour = day)) +
  geom_jitter() + 
  geom_violin(alpha = 0.5) + 
  scale_color_manual(values = day.cols) + 
  scale_fill_manual(values = day.cols) + 
  theme_classic() + 
  labs(x = "Day", y = "Myogenin ΔCq") + 
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 13))
```



## **p21 Expression**

### *Modelling*
```{r}
p21.mod <- lmer(deltacq ~ condition*day + (1|trial), data = p21)
p21.res <- simulateResiduals(p21.mod)
plot(p21.res)
```

```{r}
summary(p21.mod)
```

Rank deficient again, so dropping day MB

```{r}
p21.diff <- p21 %>%
  filter(day == "1" | day == "2")
  
p21.diff.mod <- lmer(deltacq ~ condition * day + (1 | trial), data = p21.diff)
p21.diff.res <- simulateResiduals(p21.diff.mod)
plot(p21.diff.res) # sig issues with quantile test
```

```{r}
summary(p21.diff.mod)
```

- Intercept: Represents the baseline predicted `deltacq` for 10FBS at
  Day 1.

- Condition main effects (day 1): None of the media conditions differ
  significantly from the 10FBS baseline on Day 1.

- Time main effect (day2): `deltacq` does not change significantly
  between Day 1 and Day 2 within 10FBS.

- Condition \* Day interactions: None of the interaction terms are
  significant.

```{r}
p21.noint.mod <- lmer(deltacq ~ condition + day + (1 | trial), data = p21.diff)
noint.res <- simulateResiduals(p21.noint.mod)
plot(noint.res) # no issues 
```

```{r}
summary(p21.noint.mod)
```

- Intercept: Represents the baseline predicted `deltacq` for 10FBS at
  Day 1.
  
  * Condition effects (averaged across days): No significant differences exist between conditions
  
  * Time Effect (averaged across conditions): decrease in `deltacq` on day 2 relative to day 1, but not significant (p = 0.0526) 


Extracting estimated marginal means: 
```{r}
p21.cond.emmeans <- emmeans(p21.noint.mod, pairwise ~ condition | day)
p21.cond.emmeans
```

```{r}
p21.day.emmeans <- emmeans(p21.noint.mod, pairwise ~ day | condition)
p21.day.emmeans
```


### *Visualization*
```{r}
ggplot(p21, aes(x = day, y = deltacq, fill = day, colour = day)) +
  geom_jitter() + 
  geom_violin(alpha = 0.5) + 
  scale_color_manual(values = day.cols) + 
  scale_fill_manual(values = day.cols) + 
  theme_classic() + 
  labs(x = "Day", y = "p21 ΔCq") + 
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 8))
```


## **Cell Fusion Index**

### *Modelling*
```{r}
cfi.mod <- lmer(cfi ~ condition + (1|trial), data = cfi)
cfi.res <- simulateResiduals(cfi.mod)
plot(cfi.res)
```

```{r}
summary(cfi.mod)
cfi.emm <- emmeans(cfi.mod, pairwise ~ condition)
summary(cfi.emm)
```

Significant comparisons: 

 * 10FBS - 2FBS (0.001)
 * 10FBS - 2HS (<0.001)
 * 10HS - 2FBS (<0.001)
 * 10HS - 2 HS (<0.001)
 
### *Visualization*

Colour palette for conditions (10FBS, 10HS, 2FBS, 2HS)
```{r}
cond.cols <- pnw_palette(name = "Sunset", n = 4, type = "discrete")
```

```{r}
ggplot(cfi, aes(x = condition, y = cfi, fill = condition, colour = condition)) +
  geom_jitter() + 
  geom_violin(alpha = 0.5) + 
  scale_color_manual(values = cond.cols) + 
  scale_fill_manual(values = cond.cols) + 
  theme_classic() + 
  labs(x = "Condition", y = "Cell Fusion Index (CFI)") + 
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 110))
```

Note: this x-axis can be rearranged depending on what order you want the conditions to be presented in


## **Myotube Diameter**

### *Modelling*
```{r}
diam.lmod <- lmer(avg.diam ~ condition*day + (1|trial), data = diam.clean)
diam.lres <- simulateResiduals(diam.lmod)
plot(diam.lres) # singular and quantile 
```

```{r}
diam.mod <- lmer(avg.diam ~ condition * day + (1|trial), data = diam.clean)
diam.res <- simulateResiduals(diam.mod)
plot(diam.res) # singular and quantile, with or without interaction
```

```{r}
diam.noRE <- lm(avg.diam ~ condition * day + trial, data = diam.clean)
diam.noRE.res <- simulateResiduals(diam.noRE)
plot(diam.noRE.res) # looks good 
```

```{r}
summary(diam.noRE)
```

```{r}
diam.condDay <- emmeans(diam.noRE, pairwise ~ condition | day) 
diam.condDay

diam.dayCond <- emmeans(diam.noRE, pairwise ~ day | condition)
diam.dayCond
```


### *Visualization*

```{r}
ggplot(diam, aes(x = day, y = diameter, colour = condition, fill = condition)) + 
  stat_summary(fun = mean, geom = "point", alpha = 0.8) + 
  stat_summary(fun = mean, geom = "line", aes(group = condition), linewidth = 1, lty = 2) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1, alpha = 0.8) + 
  scale_colour_manual(values = cond.cols) +
  scale_fill_manual(values = cond.cols) +
  theme_classic() + 
  labs(x = "Day", y = "Myotube Diameter (μm)", colour = "Condition", fill = "Condition") + 
  coord_cartesian(ylim = c(0, 100))
```

